---
title: "Downloading & Processing CHIRPS satellite rainfall data"
author: "brouwern@gmail.com"
date: "July 6, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction

* CHIRPS is a satellite-derived estimate of daily rainfall that has been corrected with ground station data.
* THis script downloads and processes CHIRPS data
* Functions in the curl library are used to download CHIRPS data from the ftp site.
* In this script I download data for the Carribean region and then subset for my focal region in the Dominican Republic

## References
Funk, C., P. Peterson, M. Landsfeld, D. Pedreros, J. Verdin, S. Shukla, G. Husak, J. Rowland, L. Harrison, A. Hoell, and J. Michaelsen. 2015. The climate hazards infrared precipitation with stations—a new environmental record for monitoring extremes. Scientific Data 2:150066.


## Libraries

```{r}
library(curl)
library(raster)
library(R.utils)
library(reshape2)
library(lubridate)
```


## Location of CHIRPS carib data
```{r}
ftp. <- "ftp://ftp.chg.ucsb.edu/pub/org/chg/products/CHIRPS-2.0/camer-carib_monthly/tifs/"
```




## Access list of CHIRPS files 
https://stackoverflow.com/questions/36645445/downloading-files-from-ftp-with-r

```{r}
h <- new_handle(dirlistonly=TRUE)

con <- curl(ftp., "r", h)

tbl <- read.table(con, stringsAsFactors=TRUE, fill=TRUE)
close(con)
head(tbl)
length(tbl)

```



## Test run: download 2016 data

### Identify desired files

```{r}
i. <- grep("2016",tbl[,1])
```



### Download desired file
 
```{r}
ftp.file <- paste0(ftp., tbl[i.[1],1])
fls = basename(ftp.file)
curl_fetch_disk(ftp.file[1], fls[1])
```


### Unzip file

```{r}

gunzip(fls,
       remove = FALSE)
```


### Load file


Load file as raster
```{r}
fls <- gsub(".gz","",fls)
rast <- stack(fls)
```

#### Plot data as-is

Values of "-9999" indicate missing or otherwise unusable data.
```{r}
plot(rast, maxpixels = c(500000*0.1))
summary(rast@data@values)
```

#### Remove NAs
```{r}
rast2 <- reclassify(rast,cbind(-9999,NA))

plot(rast2, maxpixels = c(500000*0.1))
```

## Download in Bulk


### Download
```{r}
ftp. <- "ftp://ftp.chg.ucsb.edu/pub/org/chg/products/CHIRPS-2.0/camer-carib_monthly/tifs/"

h <- new_handle(dirlistonly=TRUE)

con <- curl(ftp., "r", h)

tbl <- read.table(con, stringsAsFactors=TRUE, fill=TRUE)
close(con)
head(tbl)



## Identify desired files
files.i <- grep("[2][0][1][6]",tbl[,1])


dir.create("CHIRPS_gz")

## Download desired file
for(i in 1:length(files.i)){
  ftp.file <- paste0(ftp., tbl[files.i[i],1])
  fls = basename(ftp.file)
  curl_fetch_disk(ftp.file[1], fls[1],
                  path = "./CHIRPS_tiffs")
}

```


### Unzip in bulk

```{r}
#unzip
dir.create("CHIRPS_tiff")
all.gz <- list.files("./CHIRPS_gz",full.names = T)

for(i in 1:length(all.gz)){
  gunzip(all.gz[[i]],
       remove = FALSE)
}


#move
all.tif <- list.files("./CHIRPS_gz",full.names = T, pattern = "[t][i][f]$")

for(i in 1:length(all.tiff)){
  new.file <- gsub("_gz","_tiff",all.tiff[i])
  file.rename(from  = all.tiff[i],
            to = new.file)
}

```


### Load files as raster

```{r}
all.tif <- list.files("./CHIRPS_tiff",full.names = T, pattern = "[t][i][f]$")

tif.list <- vector("list", length(all.tif))

names(tif.list) <-gsub("(.*)([c][h][i][r][p][s])([-])(.)","\\2\\4",all.tif)



for(i in 1:length(all.tif)){
  rast.work <- stack(all.tif[[i]])
  rast.work <- reclassify(rast.work,cbind(-9999,NA))
  tif.list[[i]] <- rast.work
}
  

```

### Stack rasters

```{r}
tif.stack <- stack(tif.list)
```



## Process CHIRPS data

### Get outline of focal countries

```{r}
#outline of DR
DOM.shp <- getData("GADM", 
               country = c("DOM"), 
               level = 0, 
               path = tmpDir())

#outline of Haiti
HIT.shp <- getData("GADM", 
               country = c("HTI"), 
               level = 0, 
               path = tmpDir())

```


Check results
```{r}
par(mfrow = c(1,2))
plot(HIT.shp)    
plot(DOM.shp)
```




### Join DR and Haiti extents

union() joins the 2 shapes
```{r}
HISP.shp <- union(DOM.shp,HIT.shp)
```

Check results
```{r}
par(mfrow = c(1,1))
plot(HISP.shp)    

```

Plot extent over CHRIPs
```{r}
plot(tif.list$chirpsv2.0.2016.10.tif)

plot(HISP.shp,add=TRUE)

```



### Subset focal extent

Use drawExtent to draw square around focal area
```{r}
plot(tif.list$chirpsv2.0.2016.10.tif)

ext <- drawExtent()
```


Crop the stack of raster data to extent
```{r}

tif.stack.hisp <- crop(x = tif.stack, y = ext)
i<-7
plot(tif.stack.hisp[[i]],
     main = names(tif.stack.hisp[[i]]))



```


Save cropped raster
```{r}
save(tif.stack.hisp,file = "./data/data_out/temp.RData")

```

```{r}
load("./data/data_out/temp.RData")
```



## Merge CHIRPS rainfall w/ spatial locations of study sites

### Load spatial points for study site
```{r}
load("./data/data_in/pedernales_spatial_points.RData")
```


## Align spatial points to raster

```{r}
p.all <- spTransform(spdf.all,
                     crs(tif.stack.hisp[[1]]))
```

Check

```{r}
plot(tif.stack.hisp[[1]][[1]])
plot(p.all, add = TRUE)
```



## Identify cell that a point lies within
gis.stackexchange.com/questions/174925/how-to-find-the-cell-location-index-of-a-raster-using-lat-long-information



Get cell numbers
```{r}

# get cell number
##thorn-scrub site site 
id.scrub <- extract(tif.stack.hisp[[1]],
                   SpatialPoints(cbind(p.all@data[1,2],
                                       p.all@data[1,1])), 
                   cellnumbers=TRUE)[1]

##broadleaf forest
id.bl <- extract(tif.stack.hisp[[1]],
                   SpatialPoints(cbind(p.all@data[3,2],
                                       p.all@data[3,1])), 
                   cellnumbers=TRUE)[1]


## Lloyds site (Lloyd et al 2015 PeerJ)
id.Lloyd<- extract(tif.stack.hisp[[1]],
                   SpatialPoints(cbind(p.all@data[8,2],
                                       p.all@data[8,1])), 
                   cellnumbers=TRUE)[1]



```


Get rows x column location
```{r}
row.col.scrub <- rowColFromCell(tif.stack.hisp[[1]],
               id.scrub)

row.col.bl <- rowColFromCell(tif.stack.hisp[[1]],
               id.bl)


row.col.Lloyd <- rowColFromCell(tif.stack.hisp[[1]],
               id.Lloyd)
```




# Visualize cell locations
Function to visualize selected cells
```{r}
plot.cell <- function(rast, row.col, color = 2){
  plot(extent(rast, 
            row.col[1], row.col[1], 
            row.col[2],  row.col[2]), 
     add=TRUE, col=color, lwd=1)
}
```


Plot selected cells
```{r}
plot(tif.stack.hisp[[1]])
plot(p.all, add = TRUE)
plot.cell(rast = tif.stack.hisp[[1]],
          row.col = row.col.scrub)
plot.cell(rast = tif.stack.hisp[[1]],
          row.col = row.col.bl,
          color = 3)
plot.cell(rast = tif.stack.hisp[[1]],
          row.col = row.col.Lloyd,
          color = 4)
```





## Build time series

Get meta data for each layer
```{r}
#get names of each layr
layer.names <- names(tif.stack.hisp)

#extract year month etc
yr <- gsub(".*([21][01-9][01-9][01-9]).*","\\1",layer.names)
mo.no.char <- gsub(".*([01][01-9])([.][t][i][f])$","\\1",layer.names)


```


Extract data values from stacked data; "ts" is for "time series"
```{r}
ts.scrub = as.vector(tif.stack.hisp[row.col.scrub])
ts.bl = as.vector(tif.stack.hisp[row.col.bl])
ts.Lloyd = as.vector(tif.stack.hisp[row.col.Lloyd])

```


Put data from selected cells into into dataframes
```{r}
dat.scrub <- data.frame(rainfall = ts.scrub,
                        yr = yr,
                        mo.no.char = mo.no.char,
                        site = "scrub",
                        stringsAsFactors = F,
                        i = 1:length(ts.scrub))

dat.broadleaf <- data.frame(rainfall = ts.bl,
                        yr = yr,
                        mo.no.char = mo.no.char,
                        site = "broadleaf",
                        stringsAsFactors = F,
                        i = 1:length(ts.bl))


dat.Lloyd <- data.frame(rainfall = ts.Lloyd,
                        yr = yr,
                        mo.no.char = mo.no.char,
                        stringsAsFactors = F,
                        site = "Lloyd",
                        i = 1:length(ts.Lloyd))

#stack dataframes
dat <- rbind(dat.scrub,
             dat.broadleaf,
             dat.Lloyd)




```


Calculate month

```{r}
mo.no <- gsub("^[0]","",dat$mo.no.char)
dat$mo.no <- as.numeric(mo.no)

dat$mo.char <- month.abb[as.numeric(dat$mo.no)]


```






## Plot

```{r}
qplot(y = rainfall,
      x = yr,
      data = na.omit(dat))
```




## Formulate dates

Assign day as the approximate middle of each month
```{r}
dat$day <- 15
```


Build dates
```{r}
date. <- with(dat,
                 paste(day,mo.no,yr,sep = "/"))

date. <- with(dat,
                 paste(yr,mo.no.char,day,sep = "-"))

date.[length(date.)] <- NA

dat$date <- date.

```


## Calculate Julian datas

yday function is from lubridate

calculate julian where Jan 1st = 1
```{r}

dat$julian <- yday(as.Date(date.))

```



## Plot based on Julian

```{r}
qplot(y = rainfall,
      x = julian,
      data = na.omit(dat),
      group = yr,
      geom = c("point","line"),
      facets = site ~ .)
```


## Save monthly data

```{r}
write.csv(dat, 
          file= "./data/data_out/CHIRPS_rainfall_monthly_time_series.csv",
          row.names = FALSE)

```


### Reload
```{r}
dat <- read.csv("./data/data_out/CHIRPS_rainfall_monthly_time_series.csv")
```



## Summarize by season

### Define seasons

```{r}
dat$season <- "win"
dat$season <- ifelse(dat$mo.no > 3, "spr", dat$season )
dat$season <- ifelse(dat$mo.no > 6, "sum", dat$season )
dat$season <- ifelse(dat$mo.no > 9, "fal", dat$season )

dat$season <- factor(dat$season,
                     levels = c("win","spr",
                                "sum","fal"))
with(dat, table(season, mo.no))

```


Calculate seasonal summaries (Winter, spring, summer, fall)
```{r}
library(reshape2)
dat <- na.omit(dat)

## seasonal rainfall
### total rainfall by season (sum)
rain.season.sum <- dcast(data = dat,
                      yr  + site ~  season,
                      value.var = "rainfall",
                      fun.aggregate = sum)

names(rain.season.sum)[-c(1:2)] <- paste("rain.sum",
                          names(rain.season.sum)[-c(1:2)],
                          sep = ".")

i.2017 <- which(rain.season.sum$yr == 2017)
rain.season.sum[i.2017,c("rain.sum.sum","rain.sum.fal")] <- NA


## mean rainfall by season
rain.season.mean <- dcast(data = dat,
                      yr   + site ~  season,
                      value.var = "rainfall",
                      fun.aggregate = mean)

names(rain.season.mean)[-c(1:2)] <- paste("rain.mean",
                                        names(rain.season.mean)[-c(1:2)],
                                         sep = ".")
```


Yearly rainfall
```{r}

## yearly rainfall
### total rainfall by year (annual sum)
rain.yr.sum <- dcast(data = dat,
                      yr   + site ~  .,
                      value.var = "rainfall",
                      fun.aggregate = sum)
names(rain.yr.sum)[3] <- "rain.sum.yr"
rain.yr.sum[i.2017,"rain.sum.yr"] <- NA


### mean rainfall by year 
rain.yr.mean <- dcast(data = dat,
                      yr   + site ~  .,
                      value.var = "rainfall",
                      fun.aggregate = mean)
names(rain.yr.mean)[3] <- "rain.mean.yr"
rain.yr.mean[i.2017,"rain.mean.yr"] <- NA





```


2nd "Rainy season" rainfall
```{r}
cast.by.month <- dcast(data = dat,
                      yr  + site ~  mo.no,
                      value.var = "rainfall",
                      fun.aggregate = sum)

rainy.season2 <- apply(cast.by.month[,c(11,12,13)],1,sum)
rainy.season2[which(rainy.season2==0)] <- NA

```






Compile

```{r}
temp1 <- merge(rain.yr.mean,rain.yr.sum)
temp2 <- merge(rain.season.sum,rain.season.mean)

rain.ann.summary <- merge(temp1,temp2)

rain.ann.summary <- cbind(rain.ann.summary,
                          rainy.season2)

```



## Explore correlations


```{r}
library(ggplot2)
library(GGally)
rain.ann.summary2 <- rain.ann.summary

names(rain.ann.summary2) <- 
gsub("rain","R",names(rain.ann.summary2))

names(rain.ann.summary2) <- gsub("mean","mn",names(rain.ann.summary2))
names(rain.ann.summary2) <- gsub("sum","su",names(rain.ann.summary2))


i.sum <- grep("rain.sum",
              names(rain.ann.summary))
rain.ann.summary3 <- rain.ann.summary2[,-i.sum]

i.lloyd <- which(rain.ann.summary3$site == "Lloyd")

ggpairs(data = rain.ann.summary3[-i.lloyd,],
        mapping = aes(color = site), 
        columns = 3:ncol(rain.ann.summary3)) 
```



## Save annual summary

```{r}
write.csv(rain.ann.summary, file = "./data/data_out/CHIRPS_rainfall_annual_summary.csv",
          row.names = F)
```

