---
title: "rainfall_CHIRPS_Faaborg_summaries"
author: "Emily Scott"
date: "July 28, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This script works up the rainfall predictors used by Faaborg in their study in Puerto Rico.

## NAMING CONVENTION

There are one year and 6 month rain intervals (one year = "full", 6 months = "half")

Example: for measurements taken in Jan. 2017

* rain.full._: total rainfall from Jan. 2016 - Dec. 2016 (the previous year)
* rain.full.full: total rainfrom from Jan. 2015 - Dec. 2016 (the previous 2 years)
* rain.half._: total rainfall from Jan. 2016 - Jun. 2016 (1st 6 mo of prev year)
* rain.half.half: total rainfall from Jan. 2015 - Jun. 2015 & Jan. 2016 - Jun. 2016 (1st 6 months of previous 2 years)
* rain._.half: total rainfall from Jan. 2015 - Jun. 2015 (1st 6 months from 2 years ago)

## Load data
```{r}
dat <- read.csv("./data/data_out/CHIRPS_rainfall_monthly_time_series.csv")
```

## Define max and min year
## Min year is fixed at 1995 because CHIRPS data don't exist before that
```{r}
max.yr <- max(dat$yr)
min.yr <- 1995
```

## Create dataframe to store all climate data
```{r}
# Create dataframe
clim.dat <- matrix(nrow = c(3*(max.yr-min.yr-1)), ncol = 2)
clim.dat <- as.data.frame(clim.dat)

# Rename columns
colnames(clim.dat) <- c("year", "site")

# Fill 1st two columns with relevant info
for (i in 1:max.yr-min.yr-1)
{
  current.yr <- 1 + min.yr + i
  
  clim.dat$year[3*i-2] <- c(1 + min.yr + i)
  clim.dat$year[3*i-1] <- c(1 + min.yr + i)
  clim.dat$year[3*i]   <- c(1 + min.yr + i)
  
  clim.dat$site[3*i-2] <- "scrub"
  clim.dat$site[3*i-1] <- "broadleaf"
  clim.dat$site[3*i]   <- "Lloyd"
}
```

# PREV: Total rainfall in previous year
```{r}
# Create column to store one year rainfall data
clim.dat$rain.full._ <- rep(NA, dim(clim.dat)[1])

# Fill column with one year rainfall data
for (i in 1:max.yr-min.yr-1)
{
  current.yr <- 1 + min.yr + i
  
  clim.dat$rain.full._[3*i-2] <- sum(dat[which(dat$yr == current.yr-1 & dat$site == "scrub"),
                                         "rainfall"])
  clim.dat$rain.full._[3*i-1] <- sum(dat[which(dat$yr == current.yr-1 & dat$site == "broadleaf"),
                                         "rainfall"])
  clim.dat$rain.full._[3*i]   <- sum(dat[which(dat$yr == current.yr-1 & dat$site == "Lloyd"),
                                         "rainfall"])
}
```

## SUM: Sum of total rainfall of the current year and the previous year
```{r}
# Create column to store two year rainfall data
clim.dat$rain.full.full <- rep(NA, dim(clim.dat)[1])

# Fill column with two year rainfall data
for (i in 1:max.yr-min.yr-1)
{
  current.yr <- 1 + min.yr + i
  
  clim.dat$rain.full.full[3*i-2] <- sum(dat[which(dat$yr == current.yr-1 & dat$site == "scrub"| 
                                                  dat$yr == current.yr-2 & dat$site == "scrub"),
                                            "rainfall"])
  clim.dat$rain.full.full[3*i-1] <- sum(dat[which(dat$yr == current.yr-1 & dat$site == "broadleaf"| 
                                                  dat$yr == current.yr-2 & dat$site == "broadleaf"),
                                            "rainfall"])
  clim.dat$rain.full.full[3*i]   <- sum(dat[which(dat$yr == current.yr-1 & dat$site == "Lloyd"| 
                                                  dat$yr == current.yr-2 & dat$site == "Lloyd"),
                                            "rainfall"])
}
```

## FIR: Rainfall in 1st six months of previous year (January-June)
```{r}
# Create column to store 6 month rainfall data from previous year
clim.dat$rain.half._ <- rep(NA, dim(clim.dat)[1])

# Fill column with 6 month rainfall data from previous year
for (i in 1:max.yr-min.yr-1)
{
  current.yr <- 1 + min.yr + i

  clim.dat$rain.half._[3*i-2] <- sum(dat[which(dat$yr == current.yr-1 & dat$site == "scrub" & dat$mo.no %in% c(1:6)),
                                         "rainfall"])
  clim.dat$rain.half._[3*i-1] <- sum(dat[which(dat$yr == current.yr-1 & dat$site == "broadleaf" & dat$mo.no %in% c(1:6)),
                                         "rainfall"])
  clim.dat$rain.half._[3*i]   <- sum(dat[which(dat$yr == current.yr-1 & dat$site == "Lloyd" & dat$mo.no %in% c(1:6)),
                                         "rainfall"])
}
```

## F_SUM: Sum of total rainfall in 1st six months of current and previous year
```{r}
# Create column to store 6 month rainfall data from previous two years
clim.dat$rain.half.half <- rep(NA, dim(clim.dat)[1])

# Fill column with 6 month rainfall data from previous two years
for (i in 1:max.yr-min.yr-1)
{
  current.yr <- 1 + min.yr + i

  clim.dat$rain.half.half[3*i-2] <- sum(dat[which(dat$yr == current.yr-1 & dat$site == "scrub" & dat$mo.no %in% c(1:6)| 
                                                  dat$yr == current.yr-2 & dat$site == "scrub" & dat$mo.no %in% c(1:6)),
                                            "rainfall"])
  clim.dat$rain.half.half[3*i-1] <- sum(dat[which(dat$yr == current.yr-1 & dat$site == "broadleaf" & dat$mo.no %in% c(1:6)| 
                                                  dat$yr == current.yr-2 & dat$site == "broadleaf" & dat$mo.no %in% c(1:6)),
                                            "rainfall"])
  clim.dat$rain.half.half[3*i]   <- sum(dat[which(dat$yr == current.yr-1 & dat$site == "Lloyd" & dat$mo.no %in% c(1:6)| 
                                                  dat$yr == current.yr-2 & dat$site == "Lloyd" & dat$mo.no %in% c(1:6)),
                                            "rainfall"])
}
```

# F_PREV: Total rainfall from 1st six months of previous year
```{r}
# Create column to store 6 month rainfall data from 2 years ago
clim.dat$rain._.half <- rep(NA, dim(clim.dat)[1])

# Fill column with 6 month rainfall data from 2 years ago
for (i in 1:max.yr-min.yr-1)
{
  current.yr <- 1 + min.yr + i

  clim.dat$rain._.half[3*i-2] <- sum(dat[which(dat$yr == current.yr-2 & dat$site == "scrub" & dat$mo.no %in% c(1:6)),
                                         "rainfall"])
  clim.dat$rain._.half[3*i-1] <- sum(dat[which(dat$yr == current.yr-2 & dat$site == "broadleaf" & dat$mo.no %in% c(1:6)),
                                         "rainfall"])
  clim.dat$rain._.half[3*i]   <- sum(dat[which(dat$yr == current.yr-2 & dat$site == "Lloyd" & dat$mo.no %in% c(1:6)),
                                         "rainfall"])
}
```




## Save Data

```{r}
dat <- write.csv(clim.dat,
                 file = "./data/data_out/CHIRPS_rainfall_monthly_time_series_Faaborg.csv")
```

